---
title: "Equivalence Detection Model Selection"
output:
  html_notebook:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5.5
)

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("This notebook requires ggplot2.")
}
```

This notebook summarizes the equivalence-detection benchmark as a model-selection
problem for a single-call deployment: in a real run, each mutant is normally sent
to the model once. It assumes that false `EQUIVALENT` calls are the highest-risk
error, because they can hide real surviving mutants.

```{r load-data}
candidate_dirs <- c(
  "benchmarks/results",
  "results",
  "."
)
required <- c(
  "equivalence_benchmark_summary.csv",
  "equivalence_ground_truth_model_comparison.csv"
)
results_dir <- candidate_dirs[vapply(candidate_dirs, function(d) {
  all(file.exists(file.path(d, required)))
}, logical(1))][1]
if (is.na(results_dir)) {
  stop("Could not find benchmark result CSV files.")
}

summary_path <- file.path(results_dir, "equivalence_benchmark_summary.csv")
comparison_path <- file.path(results_dir, "equivalence_ground_truth_model_comparison.csv")

summary <- read.csv(summary_path, stringsAsFactors = FALSE)
comparison <- read.csv(comparison_path, stringsAsFactors = FALSE)
```

```{r compute-metrics}
aggregate_time <- aggregate(
  cbind(time_mean_s, changed_mutants, avg_dont_know) ~ model,
  summary,
  sum
)
aggregate_time_sd <- aggregate(time_sd_s ~ model, summary, function(x) {
  sqrt(sum(x^2, na.rm = TRUE))
})
names(aggregate_time_sd)[names(aggregate_time_sd) == "time_sd_s"] <- "total_time_sd_s"
aggregate_time <- merge(aggregate_time, aggregate_time_sd, by = "model")

call_rows <- do.call(rbind, lapply(seq_len(nrow(comparison)), function(i) {
  row <- comparison[i, ]
  verdicts <- strsplit(row$model_runs, ";", fixed = TRUE)[[1]]
  data.frame(
    package = row$package,
    mutant_id = row$mutant_id,
    model = row$model,
    run = seq_along(verdicts),
    model_verdict = verdicts,
    ground_truth_verdict = row$ground_truth_verdict,
    stringsAsFactors = FALSE
  )
}))
call_rows$matches_ground_truth <- call_rows$model_verdict == call_rows$ground_truth_verdict

model_rows <- lapply(sort(unique(call_rows$model)), function(model) {
  x <- call_rows[call_rows$model == model, ]
  gt_equivalent <- x$ground_truth_verdict == "EQUIVALENT"
  predicted_equivalent <- x$model_verdict == "EQUIVALENT"

  data.frame(
    model = model,
    gt_agreement_pct = round(mean(x$matches_ground_truth) * 100, 1),
    equivalent_precision_pct = if (sum(predicted_equivalent)) {
      round(sum(predicted_equivalent & gt_equivalent) / sum(predicted_equivalent) * 100, 1)
    } else {
      NA_real_
    },
    equivalent_recall_pct = round(sum(predicted_equivalent & gt_equivalent) / sum(gt_equivalent) * 100, 1),
    false_equivalent = sum(predicted_equivalent & !gt_equivalent),
    missed_equivalent = sum(!predicted_equivalent & gt_equivalent),
    predicted_equivalent = sum(predicted_equivalent),
    predicted_dont_know = sum(x$model_verdict == "DONT KNOW"),
    total_calls = nrow(x),
    stringsAsFactors = FALSE
  )
})
model_metrics <- do.call(rbind, model_rows)
model_metrics <- merge(model_metrics, aggregate_time, by = "model")
names(model_metrics)[names(model_metrics) == "time_mean_s"] <- "total_mean_time_s"
names(model_metrics)[names(model_metrics) == "changed_mutants"] <- "repeat_disagreements"
model_metrics$dont_know_total <- round(model_metrics$avg_dont_know * 3)
model_metrics$total_time_cv_pct <- round(
  model_metrics$total_time_sd_s / model_metrics$total_mean_time_s * 100,
  1
)

is_dominated <- function(i, x) {
  objectives <- c(
    "total_mean_time_s",
    "false_equivalent",
    "dont_know_total",
    "repeat_disagreements",
    "missed_equivalent"
  )
  for (j in seq_len(nrow(x))) {
    if (i == j) next
    no_worse <- all(x[j, objectives] <= x[i, objectives])
    strictly_better <- any(x[j, objectives] < x[i, objectives])
    if (no_worse && strictly_better) return(TRUE)
  }
  FALSE
}

model_metrics$pareto_front <- !vapply(seq_len(nrow(model_metrics)), is_dominated, logical(1), x = model_metrics)
model_metrics <- model_metrics[order(
  model_metrics$false_equivalent,
  -model_metrics$gt_agreement_pct,
  model_metrics$total_mean_time_s
), ]

write.csv(
  model_metrics,
  file.path(results_dir, "equivalence_model_selection_metrics.csv"),
  row.names = FALSE
)
```

## Model Selection Table

Counts are aggregated across the three benchmark packages and use each repeated
model call as one deployment-like observation. There are 225 single-call verdicts
per model: 75 selected mutants times 3 repeats.

```{r model-table}
display_table <- model_metrics[, c(
  "model",
  "total_mean_time_s",
  "total_time_sd_s",
  "total_time_cv_pct",
  "gt_agreement_pct",
  "equivalent_precision_pct",
  "equivalent_recall_pct",
  "false_equivalent",
  "predicted_equivalent",
  "dont_know_total",
  "repeat_disagreements",
  "pareto_front"
)]
names(display_table) <- c(
  "Model",
  "Total mean time (s)",
  "Total SD (s)",
  "CV (%)",
  "GT agreement (%)",
  "E precision (%)",
  "E recall (%)",
  "False E",
  "E calls",
  "DONT KNOW",
  "Repeat disagreements",
  "Pareto front"
)
knitr::kable(display_table, row.names = FALSE, digits = 1)
```

## Pareto-Style Utility Plot

This plot emphasizes the main tradeoff for a single-call workflow: runtime versus
ability to find ground-truth equivalent mutants. Color encodes the high-risk
single-call error count (`false_equivalent`), and point size encodes repeat
instability.

```{r pareto-plot}
library(ggplot2)

model_metrics$label_side <- ifelse(
  model_metrics$total_mean_time_s < stats::median(model_metrics$total_mean_time_s),
  -0.08,
  1.08
)
model_metrics$label_hjust <- ifelse(model_metrics$label_side < 0, 1, 0)

ggplot(
  model_metrics,
  aes(
    x = total_mean_time_s,
    y = equivalent_recall_pct,
    colour = false_equivalent,
    size = repeat_disagreements,
    shape = pareto_front,
    label = model
  )
) +
  geom_point(alpha = 0.9, stroke = 1.1) +
  geom_text(
    aes(y = equivalent_recall_pct + label_side, hjust = label_hjust),
    size = 3,
    check_overlap = TRUE,
    colour = "grey15"
  ) +
  scale_x_log10(
    breaks = c(30, 50, 100, 200, 500, 1000, 2000),
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    limits = c(0, 105),
    expand = ggplot2::expansion(mult = c(0.03, 0.08))
  ) +
  scale_colour_gradient(
    low = "#2c7bb6",
    high = "#d7191c",
    breaks = sort(unique(model_metrics$false_equivalent))
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) +
  labs(
    x = "Total mean time across packages (seconds, log scale)",
    y = "Equivalent-mutant recall vs provisional ground truth (%)",
    colour = "False E",
    size = "Repeat\ndisagreements",
    shape = "Pareto front",
    title = "Single-call tradeoff: speed, equivalent recall, and false-equivalent risk"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.margin = margin(10, 42, 10, 62)
  )
```

## Safety Plot

This second view puts false `EQUIVALENT` calls directly on the y-axis. The best
single-call models are near the lower-left; larger points find more of the
ground-truth equivalent mutants.

```{r safety-plot}
ggplot(
  model_metrics,
  aes(
    x = total_mean_time_s,
    y = false_equivalent,
    colour = gt_agreement_pct,
    size = equivalent_recall_pct,
    shape = pareto_front,
    label = model
  )
) +
  geom_point(alpha = 0.9, stroke = 1.1) +
  geom_text(
    aes(y = false_equivalent + label_side / 2, hjust = label_hjust),
    size = 3,
    check_overlap = TRUE,
    colour = "grey15"
  ) +
  scale_x_log10(
    breaks = c(30, 50, 100, 200, 500, 1000, 2000),
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    breaks = seq(0, max(model_metrics$false_equivalent), by = 1),
    limits = c(-0.25, max(model_metrics$false_equivalent) + 0.5),
    expand = ggplot2::expansion(mult = c(0.02, 0.08))
  ) +
  scale_colour_gradient(low = "#d9ef8b", high = "#1a9850") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) +
  labs(
    x = "Total mean time across packages (seconds, log scale)",
    y = "False EQUIVALENT calls",
    colour = "GT agreement (%)",
    size = "E recall (%)",
    shape = "Pareto front",
    title = "Single-call safety view: avoid false equivalent classifications"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.margin = margin(10, 42, 10, 62)
  )
```

## Practical Reading

For a single-call workflow, no model in this benchmark is clean enough to make
fully automatic removal decisions if false `EQUIVALENT` calls are unacceptable:
every model made at least one false `EQUIVALENT` call across the repeated
single-call observations.

`deepseek-v4-pro` is the strongest default for assisted triage: it has high
agreement with the provisional ground truth, comparatively high precision for
`EQUIVALENT` calls, low uncertainty, and reasonable runtime. `qwen3.5-122b` is
the fast triage point, but it misses most equivalent mutants and has more
uncertainty. Higher-recall models such as `gpt-oss-120b` find more equivalent
mutants but make more false `EQUIVALENT` calls, so their positive verdicts should
be manually checked or confirmed by a second model.

The repeat-disagreement column is not a voting metric here. It is a stability
diagnostic: it estimates how likely the model is to give a different answer if
the same mutant is submitted again.
